Follow-up Samples

### 2017-09-15 submission/manuscript_R2.tex
revert back to microbiome format

\textbf{Background.} Despite growing evidence that members of the gut
microbiota can drive tumorigenesis, little is known about what happens
to it after treatment for an adenoma or carcinoma. This study tests
whether treatment alters the abundance of bacterial populations
associated with disease versus those with a normal colon by sequencing
the 16S rRNA genes in feces of 67 individuals before and after treatment
(adenoma n = 22), advanced adenoma n = 19), and carcinoma n = 26).

\textbf{Results.} The communities from patients with carcinomas changed
significantly more than those with adenoma following treatment (P-value
\textless{} 0.001). Although Random Forest models classified pre and
post-treatment samples for each diagnosis group, there was little
overlap between the OTUs that were indicative of each treatment.
Training Random Forest models on a larger cohort that contained
individuals with normal colons, adenomas, advanced adenomas, and
carcinomas we assessed the positive probability of having one of three
diagnoses and whether the pre and post-treatment samples were classified
the same. Patients who had carcinomas changed towards a microbial milieu
that resembles the normal colon after treatment (P-value \textless{}
0.05). There was no significant difference in the microbiota of
individuals treated with surgery alone and those treated with
chemotherapy or chemotherapy and radiation (P-value \textgreater{}
0.05).

\textbf{Conclusions.} It was difficult to identify significant
differences between pre and post-treatment samples from patients with
adenoma and advanced adenoma. However, for carcinomas, not only were
there large changes in pre versus post-treatment samples, but these
changes were towards a more normal microbiota.

@@

Colorectal cancer (CRC) is the third most common cause of cancer deaths
in the United States (1, 2). Disease mortality has significantly
decreased, predominately due to improvements in screening (2). Despite
these improvements, there are still approximately 50,000 CRC-related
deaths per year in the United States (1). Current estimates indicate
that 20-30\% of those who undergo treatment will experience recurrence
and 35\% of all patients will die within five years (3--5).
Identification of methods to assess patients' risk of recurrence is of
great importance to reduce mortality and healthcare costs.

There is growing evidence that the gut microbiota is involved in the
progression of CRC. Mouse-based studies have identified populations of
\emph{Bacteroides fragilis}, \emph{Escherichia coli}, and
\emph{Fusobacterium nucleatum} that alter disease progression (6--10).
Furthermore, studies that shift the structure of the microbiota through
the use of antibiotics or inoculation of germ free mice with human feces
have shown that varying community compositions can result in varied
tumor burden (11--13). Collectively, these studies support the
hypothesis that the microbiota can alter the amount of inflammation in
the colon and with it the rate of tumorigenesis (14).

Building upon this evidence, several human studies have identified
unique signatures of colonic lesions (15--20). One line of research has
identified community-level differences between those bacteria that are
found on and adjacent to colonic lesions and have supported a role for
\emph{Bacteroides fragilis}, \emph{Escherichia coli}, and
\emph{Fusobacterium nucleatum} in tumorigenesis (21--23). Others have
proposed feces-based biomarkers that could be used to diagnose the
presence of colonic adenomas and carcinomas (24--26). These studies have
associated \emph{Fusobacterium nucleatum} and other oral pathogens with
colonic lesions (adenoma, advanced adenoma, and carcinoma). They have
also noted that the loss of bacteria generally thought to produce short
chain fatty acids, which can suppress inflammation, is associated with
colonic lesions. This suggests that gut bacteria have a role in
tumorigenesis with potential as useful biomarkers for aiding in the
early detection of disease (21--26).

@@

\textbf{\emph{Post-treatment samples from patients with carcinoma more
closely resemble those of a normal colon.}} Next, we determined whether
treatment changed the microbiota in a way that the post-treatment
communities resembled that of patients with normal colons. To test this,
we used an expanded cohort of 423 individuals that were diagnosed under
the same protocol as having normal colons or colons with adenoma,
advanced adenoma, or carcinoma {[}Table 2{]}. We then constructed Random
Forest models to classify the study samples, with the 3 diagnosis groups
(adenoma, advanced adenoma, or carcinoma), or having a normal colon. The
models performed moderately with CRC being the best (adenoma AUC range =
0.50 - 0.62, advanced adenoma AUC range = 0.53 - 0.67, carcinoma AUC
range = 0.71 - 0.82; Figure S1). The OTUs that were in the top 10\% of
importance for the adenoma and advanced adenoma models largely
overlapped and those OTUs that were used to classify the carcinoma
samples were largely distinct from those of the other two models
{[}Figure 3A{]}. Among the OTUs that were shared across the three models
were those populations generally considered beneficial to their host
(e.g. \emph{Faecalibacterium}, \emph{Lachnospiraceae},
\emph{Bacteroides}, \emph{Dorea}, \emph{Anaerostipes}, and
\emph{Roseburia}) {[}Figures 3B{]}. Although many of important OTUs in
the top 10\% were also included in the model differentiating between
patients with normal colons and those with carcinoma, this model also
included OTUs affiliated with populations that have previously been
associated with carcinoma (\emph{Fusobacterium}, \emph{Porphyromonas},
\emph{Parvimonas}) (24--26) {[}Figure S2{]} with some individuals
showing a marked decrease in relative abundance {[}Figure S3{]}.
Finally, we applied these three models to the pre and post-treatment
samples for each diagnosis group and quantified the change in the
positive probability of the model. A decrease in the positive
probability would indicate that the microbiota more closely resembled
that of a patient with a normal colon. There was no significant change
in the positive probability for the adenoma or advanced adenoma groups
(P-value \textgreater{} 0.05) {[}Figure 4{]}. The positive probability
for the pre and post-treatment samples from patients diagnosed with
carcinoma significantly decreased with treatment, suggesting a shift
toward a normal microbiota for most individuals (P-value = 0.001). Only,
7 of the 26 patients (26.92\%) who were diagnosed with a carcinoma had a
higher positive probability after treatment; one of those was
re-diagnosed with carcinoma on the follow up visit. These results
indicate that, although there were changes in the microbiota associated
with treatment, those experienced by patients with carcinoma after
treatment yielded gut bacterial communities of greater similarity to
that of a normal colon.

@@

There was a subset of patients (7 of the 26 with carcinomas) who
demonstrated an elevated probability of carcinoma after treatment. This
may reflect an elevated risk of recurrence. The 26.92\% prevalence of
increased carcinoma probability from our study is within the expected
rate of recurrence (20-30\% (3, 4)). We hypothesized that these
individuals may have had more severe tumors; however, the tumor severity
of these 7 individuals (1 with Stage I, 3 with Stage II, and 3 with
Stage III) was similar to the distribution observed among the other 19
patients. We also hypothesized that we may have sampled these patients
later than the rest and their communities may have reverted to a
carcinoma-associated state; however, there was not a statistically
significant difference in the length of time between sample collection
among those whose probabilities increased (331 (246 - 358) days) or
decreased (364 (301 - 434) days) (Wilcoxon Test; P-value = 0.39) (all
days data displayed as median (IQR)). Finally, it is possible that these
patients may not have responded to treatment as well as the other 19
patients diagnosed with carcinoma and so the microbiota may not have
been impacted the same way. Again, further studies looking at the role
of the microbiota in recurrence are needed to understand the dynamics
following treatment.

@@

\textbf{\emph{Study Design and Patient Sampling.}} Sampling and design
have been previously reported in Baxter, et al (24). Briefly, samples
were stored on ice for at least 24h before freezing. Although we cannot
exclude that this sampling protocol may have impacted the gut microbiota
composition all samples were subjected to the same methodology. Study
exclusion involved those who had already undergone surgery, radiation,
or chemotherapy, had colorectal cancer before a baseline fecal sample
could be obtained, had IBD, a known hereditary non-polyposis colorectal
cancer, or familial adenomatous polyposis. Samples used to build the
models for prediction were collected either prior to a colonoscopy or
between one and two weeks after initial colonoscopy. The bacterial
community has been shown to normalize back to a pre-colonoscopy
community within this time period (27). Our study cohort consisted of 67
individuals with an initial sample as described and a follow up sample
obtained between 188 - 546 days after treatment of lesion {[}Table 1{]}.
Patients were diagnosed by colonoscopic examination and
histopathological review of any biopsies taken. Patients were classified
as having advanced adenoma if they had an adenoma greater than 1 cm,
more than three adenomas of any size, or an adenoma with villous
histology. This study was approved by the University of Michigan
Institutional Review Board. All study participants provided informed
consent and the study itself conformed to the guidelines set out by the
Helsinki Declaration.

@@

\textbf{\emph{16S rRNA Gene Sequencing.}} Sequencing was completed as
described by Kozich, et al. (28). DNA extraction used the 96-well Soil
DNA isolation kit (MO BIO Laboratories) and an epMotion 5075 automated
pipetting system (Eppendorf). The V4 variable region was amplified and
the resulting product was split between four sequencing runs with
normal, adenoma, and carcinoma evenly represented on each run. Each
group was randomly assigned to avoid biases based on sample collection
location. The pre and post-treatment samples were sequenced on the same
run.

\textbf{\emph{Sequence Processing.}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences and has been
previously described (28). The general workflow using mothur included
merging paired-end reads into contigs, filtering for low quality
contigs, aligning to the SILVA database (29), screening for chimeras
using UCHIME (30), classifying with a naive Bayesian classifier using
the Ribosomal Database Project (RDP)(31), and clustered into Operational
Taxonomic Units (OTUs) using a 97\% similarity cutoff with an average
neighbor clustering algorithm (32). The number of sequences for each
sample was rarefied to 10523 to minimize the impacts of uneven sampling.

\textbf{\emph{Model Building.}} The Random Forest (33) algorithm was
used to create the three models used to classify pre and post-treatment
samples by diagnosis (adenoma, advanced adenoma, or carcinoma). The
total number of individuals in the pre versus post-treatment models was
67 individuals. There were a total of 22 individuals in the pre versus
post-treatment adenoma model, 19 individuals in the pre versus
post-treatment advanced adenoma model, and 26 individuals in the pre
versus post-treatment carcinoma model {[}Table 1{]}.

Similarly, the Random Forest (33) algorithm was also used to create the
three models used to classify normal versus diagnosis. These samples
were obtained using the same methodology as described earlier in this
section. All samples used for this component of model training were from
pre-treatment samples only. The total number of individuals in the
normal versus diagnosis models was 423 individuals {[}Table 2{]}. There
were a total of 239 individuals in the normal versus adenoma model, 262
individuals in the normal versus advanced adenoma model, and 266
individuals in the normal versus carcinoma model {[}Table 2{]}.

All models included only OTU data obtained from 16S rRNA sequencing and
were processed and cleaned using the R package caret (v6.0.76).
Optimization of the mtry hyper-parameter involved making 100 different
80/20 (train/test) splits of the data where the same proportion was
present within both the whole data set and the 80/20 split. For each of
the different splits, 20 repeated 10-fold cross validation was performed
on the 80\% component to optimize the mtry hyper-parameter by maximizing
the AUC (Area Under the Curve of the Receiver Operator Characteristic).
The resulting model was then tested on the hold out data obtained from
the 20\% component. For adenoma pre versus post-treatment models the
optimized mtrys were 2 (n = 42), 42 (n = 26), and 914 (n = 32) for
adenoma. For advanced adenoma the mtrys were 2 (n = 26), 52 (n = 14),
and 1353 (n = 60). For carcinoma the mtrys were 2 (n = 36), 41 (n = 54),
and 863 (n = 10). For the normal versus diagnosis models the optimized
mtrys were (n = NA), (n = NA), and (n = NA) for adenoma. For advanced
adenoma (n = NA), (n = NA), and (n = NA). For carcinoma (n = NA), (n =
NA), and (n = NA). The hyper-parameter, mtry, defines the number of
variables to investigate at each split before a new division of the data
was created with the Random Forest model (33).

@@

\textbf{\emph{Statistical Analysis.}} The R software package (v3.4.1)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA (34) in the vegan package
(v2.4.3). Comparisons between probabilities as well as overall
differences in the median relative abundance of each OTU between pre and
post-treatment samples utilized a paired Wilcoxon ranked sum test.
Comparisons between different treatment for lesions utilized a Kruskal
Wallis test. Where multiple comparison testing was appropriate, a
Benjamini-Hochberg (BH) correction was applied (35) and a corrected
P-value of less than 0.05 was considered significant. The P-values
reported are those that were BH corrected. Model rank importance was
determined by obtaining the median MDA from the 100, 20 repeated 10-fold
cross validation and then ranking from largest to smallest MDA.

@@

\hypertarget{refs}{}
\hypertarget{ref-siegel_cancer_2016}{}
1. Siegel RL, Miller KD, Jemal A. 2016. Cancer statistics, 2016. CA: a
cancer journal for clinicians 66:7--30.

\hypertarget{ref-haggar_colorectal_2009}{}
2. Haggar FA, Boushey RP. 2009. Colorectal cancer epidemiology:
Incidence, mortality, survival, and risk factors. Clinics in Colon and
Rectal Surgery 22:191--197.

\hypertarget{ref-hellinger_reoperation_2006}{}
3. Hellinger MD, Santiago CA. 2006. Reoperation for recurrent colorectal
cancer. Clinics in Colon and Rectal Surgery 19:228--236.

\hypertarget{ref-ryuk_predictive_2014}{}
4. Ryuk JP, Choi G-S, Park JS, Kim HJ, Park SY, Yoon GS, Jun SH, Kwon
YC. 2014. Predictive factors and the prognosis of recurrence of
colorectal cancer within 2 years after curative resection. Annals of
Surgical Treatment and Research 86:143--151.

\hypertarget{ref-national_cancer_institute_seer_nodate}{}
5. Institute NC. SEER Cancer Stat Facts: Colon and Rectum Cancer.

\hypertarget{ref-goodwin_polyamine_2011}{}
6. Goodwin AC, Destefano Shields CE, Wu S, Huso DL, Wu X, Murray-Stewart
TR, Hacker-Prietz A, Rabizadeh S, Woster PM, Sears CL, Casero RA. 2011.
Polyamine catabolism contributes to enterotoxigenic Bacteroides
fragilis-induced colon tumorigenesis. Proceedings of the National
Academy of Sciences of the United States of America 108:15354--15359.

\hypertarget{ref-abed_fap2_2016}{}
7. Abed J, Emgård JEM, Zamir G, Faroja M, Almogy G, Grenov A, Sol A,
Naor R, Pikarsky E, Atlan KA, Mellul A, Chaushu S, Manson AL, Earl AM,
Ou N, Brennan CA, Garrett WS, Bachrach G. 2016. Fap2 Mediates
Fusobacterium nucleatum Colorectal Adenocarcinoma Enrichment by Binding
to Tumor-Expressed Gal-GalNAc. Cell Host \& Microbe 20:215--225.

\hypertarget{ref-arthur_microbial_2014}{}
8. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
McCafferty J, Fodor AA, Jobin C. 2014. Microbial genomic analysis
reveals the essential role of inflammation in bacteria-induced
colorectal cancer. Nature Communications 5:4724.

\hypertarget{ref-kostic_fusobacterium_2013}{}
9. Kostic AD, Chun E, Robertson L, Glickman JN, Gallini CA, Michaud M,
Clancy TE, Chung DC, Lochhead P, Hold GL, El-Omar EM, Brenner D, Fuchs
CS, Meyerson M, Garrett WS. 2013. Fusobacterium nucleatum potentiates
intestinal tumorigenesis and modulates the tumor-immune
microenvironment. Cell Host \& Microbe 14:207--215.

\hypertarget{ref-wu_human_2009}{}
10. Wu S, Rhee K-J, Albesiano E, Rabizadeh S, Wu X, Yen H-R, Huso DL,
Brancati FL, Wick E, McAllister F, Housseau F, Pardoll DM, Sears CL.
2009. A human colonic commensal promotes colon tumorigenesis via
activation of T helper type 17 T cell responses. Nature Medicine
15:1016--1022.

\hypertarget{ref-zackular_manipulation_2016}{}
11. Zackular JP, Baxter NT, Chen GY, Schloss PD. 2016. Manipulation of
the Gut Microbiota Reveals Role in Colon Tumorigenesis. mSphere 1.

\hypertarget{ref-zackular_gut_2013}{}
12. Zackular JP, Baxter NT, Iverson KD, Sadler WD, Petrosino JF, Chen
GY, Schloss PD. 2013. The gut microbiome modulates colon tumorigenesis.
mBio 4:e00692--00613.

\hypertarget{ref-baxter_structure_2014}{}
13. Baxter NT, Zackular JP, Chen GY, Schloss PD. 2014. Structure of the
gut microbiome following colonization with human feces determines
colonic tumor burden. Microbiome 2:20.

\hypertarget{ref-flynn_metabolic_2016}{}
14. Flynn KJ, Baxter NT, Schloss PD. 2016. Metabolic and Community
Synergy of Oral Bacteria in Colorectal Cancer. mSphere 1.

\hypertarget{ref-wang_structural_2012}{}
15. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, Jia W, Cai S, Zhao L.
2012. Structural segregation of gut microbiota between colorectal cancer
patients and healthy volunteers. The ISME journal 6:320--329.

>>>>

\textbf{Background.} Colorectal cancer is a worldwide health problem.
Despite growing evidence that members of the gut microbiota can drive
tumorigenesis, little is known about what happens to it after treatment
for an adenoma or carcinoma. This study tested the hypothesis that
treatment for adenoma or carcinoma alters the abundance of bacterial
populations associated with disease to those associated with a normal
colon. We tested this hypothesis by sequencing the 16S rRNA genes in the
feces of 67 individuals before and after treatment for adenoma (N = 22),
advanced adenoma (N = 19), and carcinoma (N = 26).

\textbf{Results.} There were small changes to the bacterial community
associated with adenoma or advanced adenoma and large changes associated
with carcinoma. The communities from patients with carcinomas changed
significantly more than those with adenoma following treatment (P-value
\textless{} 0.001). Although treatment was associated with intrapersonal
changes, the change in the abundance of individual OTUs in response to
treatment was not consistent within diagnosis groups (P-value
\textgreater{} 0.05). Because the distribution of OTUs across patients
and diagnosis groups was irregular, we used the Random Forest machine
learning algorithm to identify groups of OTUs that could be used to
classify pre and post-treatment samples for each of the diagnosis
groups. Although the three models could differentiate between the pre
and post-treatment samples, there was little overlap between the OTUs
that were indicative of each treatment. To determine whether individuals
who underwent treatment were more likely to have OTUs associated with
normal colons we used a larger cohort that contained individuals with
normal colons and those with adenomas, advanced adenomas, and
carcinomas. We again built Random Forest models and measured the change
in the positive probability of having one of the three diagnoses to
assess whether the post-treatment samples received the same
classification as the pre-treatment samples. Samples from patients who
had carcinomas changed towards a microbial milieu that resembles the
normal colon after treatment (P-value \textless{} 0.05). Finally, we
were unable to detect any significant differences in the microbiota of
individuals treated with surgery alone and those treated with
chemotherapy or chemotherapy and radiation (P-value \textgreater{}
0.05).

\textbf{Conclusions.} By better understanding the response of the
microbiota to treatment for adenomas and carcinomas, it is likely that
biomarkers will eventually be validated that can be used to quantify the
risk of recurrence and the likelihood of survival. Although it was
difficult to identify significant differences between pre and
post-treatment samples from patients with adenoma and advanced adenoma,
this was not the case for carcinomas. Not only were there large changes
in pre versus post-treatment samples for those with carcinoma, but these
changes were towards a more normal microbiota.

\subsubsection{Keywords}\label{keywords}

microbiota; colorectal cancer; polyps; treatment; risk factor.

@@

Colorectal cancer (CRC) is the third most common cause of cancer deaths
in the United States {[}1,2{]}. Disease mortality has significantly
decreased, predominately due to improvements in screening {[}2{]}.
Despite these improvements, there are still approximately 50,000
CRC-related deaths per year in the United States {[}1{]}. Current
estimates indicate that 20-30\% of those who undergo treatment will
experience recurrence and 35\% of all patients will die within five
years {[}3--5{]}. Identification of methods to assess patients' risk of
recurrence is of great importance to reduce mortality and healthcare
costs.

There is growing evidence that the gut microbiota is involved in the
progression of CRC. Mouse-based studies have identified populations of
\emph{Bacteroides fragilis}, \emph{Escherichia coli}, and
\emph{Fusobacterium nucleatum} that alter disease progression
{[}6--10{]}. Furthermore, studies that shift the structure of the
microbiota through the use of antibiotics or inoculation of germ free
mice with human feces have shown that varying community compositions can
result in varied tumor burden {[}11--13{]}. Collectively, these studies
support the hypothesis that the microbiota can alter the amount of
inflammation in the colon and with it the rate of tumorigenesis
{[}14{]}.

Building upon this evidence, several human studies have identified
unique signatures of colonic lesions {[}15--20{]}. One line of research
has identified community-level differences between those bacteria that
are found on and adjacent to colonic lesions and have supported a role
for \emph{Bacteroides fragilis}, \emph{Escherichia coli}, and
\emph{Fusobacterium nucleatum} in tumorigenesis {[}21--23{]}. Others
have proposed feces-based biomarkers that could be used to diagnose the
presence of colonic adenomas and carcinomas {[}24--26{]}. These studies
have associated \emph{Fusobacterium nucleatum} and other oral pathogens
with colonic lesions (adenoma, advanced adenoma, and carcinoma). They
have also noted that the loss of bacteria generally thought to produce
short chain fatty acids, which can suppress inflammation, is associated
with colonic lesions. This suggests that gut bacteria have a role in
tumorigenesis with potential as useful biomarkers for aiding in the
early detection of disease {[}21--26{]}.

@@

\textbf{\emph{Post-treatment samples from patients with carcinoma more
closely resemble those of a normal colon.}} Next, we determined whether
treatment changed the microbiota in a way that the post-treatment
communities resembled that of patients with normal colons. To test this,
we used an expanded cohort of 423 individuals that were diagnosed under
the same protocol as having normal colons or colons with adenoma,
advanced adenoma, or carcinoma {[}Table 2{]}. We then constructed Random
Forest models to classify the study samples, with the 3 diagnosis groups
(adenoma, advanced adenoma, or carcinoma), or having a normal colon. The
models performed moderately with CRC being the best (adenoma AUC range =
0.50 - 0.62, advanced adenoma AUC range = 0.53 - 0.67, carcinoma AUC
range = 0.71 - 0.82; Figure S1). The OTUs that were in the top 10\% of
importance for the adenoma and advanced adenoma models largely
overlapped and those OTUs that were used to classify the carcinoma
samples were largely distinct from those of the other two models
{[}Figure 3A{]}. Among the OTUs that were shared across the three models
were those populations generally considered beneficial to their host
(e.g. \emph{Faecalibacterium}, \emph{Lachnospiraceae},
\emph{Bacteroides}, \emph{Dorea}, \emph{Anaerostipes}, and
\emph{Roseburia}) {[}Figures 3B{]}. Although many of important OTUs in
the top 10\% were also included in the model differentiating between
patients with normal colons and those with carcinoma, this model also
included OTUs affiliated with populations that have previously been
associated with carcinoma (\emph{Fusobacterium}, \emph{Porphyromonas},
\emph{Parvimonas}) {[}24--26{]} {[}Figure S2{]} with some individuals
showing a marked decrease in relative abundance {[}Figure S3{]}.
Finally, we applied these three models to the pre and post-treatment
samples for each diagnosis group and quantified the change in the
positive probability of the model. A decrease in the positive
probability would indicate that the microbiota more closely resembled
that of a patient with a normal colon. There was no significant change
in the positive probability for the adenoma or advanced adenoma groups
(P-value \textgreater{} 0.05) {[}Figure 4{]}. The positive probability
for the pre and post-treatment samples from patients diagnosed with
carcinoma significantly decreased with treatment, suggesting a shift
toward a normal microbiota for most individuals (P-value = 0.001). Only,
7 of the 26 patients (26.92\%) who were diagnosed with a carcinoma had a
higher positive probability after treatment; one of those was
re-diagnosed with carcinoma on the follow up visit. These results
indicate that, although there were changes in the microbiota associated
with treatment, those experienced by patients with carcinoma after
treatment yielded gut bacterial communities of greater similarity to
that of a normal colon.

@@

There was a subset of patients (7 of the 26 with carcinomas) who
demonstrated an elevated probability of carcinoma after treatment. This
may reflect an elevated risk of recurrence. The 26.92\% prevalence of
increased carcinoma probability from our study is within the expected
rate of recurrence (20-30\% {[}3,4{]}). We hypothesized that these
individuals may have had more severe tumors; however, the tumor severity
of these 7 individuals (1 with Stage I, 3 with Stage II, and 3 with
Stage III) was similar to the distribution observed among the other 19
patients. We also hypothesized that we may have sampled these patients
later than the rest and their communities may have reverted to a
carcinoma-associated state; however, there was not a statistically
significant difference in the length of time between sample collection
among those whose probabilities increased (331 (246 - 358) days) or
decreased (364 (301 - 434) days) (Wilcoxon Test; P-value = 0.39) (all
days data displayed as median (IQR)). Finally, it is possible that these
patients may not have responded to treatment as well as the other 19
patients diagnosed with carcinoma and so the microbiota may not have
been impacted the same way. Again, further studies looking at the role
of the microbiota in recurrence are needed to understand the dynamics
following treatment.

@@

\textbf{\emph{Study Design and Patient Sampling.}} Sampling and design
have been previously reported in Baxter, et al {[}24{]}. Briefly,
samples were stored on ice for at least 24h before freezing. Although we
cannot exclude that this sampling protocol may have impacted the gut
microbiota composition all samples were subjected to the same
methodology. Study exclusion involved those who had already undergone
surgery, radiation, or chemotherapy, had colorectal cancer before a
baseline fecal sample could be obtained, had IBD, a known hereditary
non-polyposis colorectal cancer, or familial adenomatous polyposis.
Samples used to build the models for prediction were collected either
prior to a colonoscopy or between one and two weeks after initial
colonoscopy. The bacterial community has been shown to normalize back to
a pre-colonoscopy community within this time period {[}27{]}. Our study
cohort consisted of 67 individuals with an initial sample as described
and a follow up sample obtained between 188 - 546 days after treatment
of lesion {[}Table 1{]}. Patients were diagnosed by colonoscopic
examination and histopathological review of any biopsies taken. Patients
were classified as having advanced adenoma if they had an adenoma
greater than 1 cm, more than three adenomas of any size, or an adenoma
with villous histology. This study was approved by the University of
Michigan Institutional Review Board. All study participants provided
informed consent and the study itself conformed to the guidelines set
out by the Helsinki Declaration.

@@

\textbf{\emph{16S rRNA Gene Sequencing.}} Sequencing was completed as
described by Kozich, et al. {[}28{]}. DNA extraction used the 96-well
Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
automated pipetting system (Eppendorf). The V4 variable region was
amplified and the resulting product was split between four sequencing
runs with normal, adenoma, and carcinoma evenly represented on each run.
Each group was randomly assigned to avoid biases based on sample
collection location. The pre and post-treatment samples were sequenced
on the same run.

\textbf{\emph{Sequence Processing.}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences and has been
previously described {[}28{]}. The general workflow using mothur
included merging paired-end reads into contigs, filtering for low
quality contigs, aligning to the SILVA database {[}29{]}, screening for
chimeras using UCHIME {[}30{]}, classifying with a naive Bayesian
classifier using the Ribosomal Database Project (RDP){[}31{]}, and
clustered into Operational Taxonomic Units (OTUs) using a 97\%
similarity cutoff with an average neighbor clustering algorithm
{[}32{]}. The number of sequences for each sample was rarefied to 10523
to minimize the impacts of uneven sampling.

\textbf{\emph{Model Building.}} The Random Forest {[}33{]} algorithm was
used to create the three models used to classify pre and post-treatment
samples by diagnosis (adenoma, advanced adenoma, or carcinoma). The
total number of individuals in the pre versus post-treatment models was
67 individuals. There were a total of 22 individuals in the pre versus
post-treatment adenoma model, 19 individuals in the pre versus
post-treatment advanced adenoma model, and 26 individuals in the pre
versus post-treatment carcinoma model {[}Table 1{]}.

Similarly, the Random Forest {[}33{]} algorithm was also used to create
the three models used to classify normal versus diagnosis. These samples
were obtained using the same methodology as described earlier in this
section. All samples used for this component of model training were from
pre-treatment samples only. The total number of individuals in the
normal versus diagnosis models was 423 individuals {[}Table 2{]}. There
were a total of 239 individuals in the normal versus adenoma model, 262
individuals in the normal versus advanced adenoma model, and 266
individuals in the normal versus carcinoma model {[}Table 2{]}.

All models included only OTU data obtained from 16S rRNA sequencing and
were processed and cleaned using the R package caret (v6.0.76).
Optimization of the mtry hyper-parameter involved making 100 different
80/20 (train/test) splits of the data where the same proportion was
present within both the whole data set and the 80/20 split. For each of
the different splits, 20 repeated 10-fold cross validation was performed
on the 80\% component to optimize the mtry hyper-parameter by maximizing
the AUC (Area Under the Curve of the Receiver Operator Characteristic).
The resulting model was then tested on the hold out data obtained from
the 20\% component. For adenoma pre versus post-treatment models the
optimized mtrys were 2 (n = 42), 42 (n = 26), and 914 (n = 32) for
adenoma. For advanced adenoma the mtrys were 2 (n = 26), 52 (n = 14),
and 1353 (n = 60). For carcinoma the mtrys were 2 (n = 36), 41 (n = 54),
and 863 (n = 10). For the normal versus diagnosis models the optimized
mtrys were 2 (n = 16), 37 (n = 44), and 707 (n = 40) for adenoma. For
advanced adenoma 2 (n = 33), 37 (n = 50), and 696 (n = 17). For
carcinoma 36 (n = 96) and 649 (n = 4). The hyper-parameter, mtry,
defines the number of variables to investigate at each split before a
new division of the data was created with the Random Forest model
{[}33{]}.

@@

\textbf{\emph{Statistical Analysis.}} The R software package (v3.4.1)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}34{]} in the vegan package
(v2.4.3). Comparisons between probabilities as well as overall
differences in the median relative abundance of each OTU between pre and
post-treatment samples utilized a paired Wilcoxon ranked sum test.
Comparisons between different treatment for lesions utilized a Kruskal
Wallis test. Where multiple comparison testing was appropriate, a
Benjamini-Hochberg (BH) correction was applied {[}35{]} and a corrected
P-value of less than 0.05 was considered significant. The P-values
reported are those that were BH corrected. Model rank importance was
determined by obtaining the median MDA from the 100, 20 repeated 10-fold
cross validation and then ranking from largest to smallest MDA.

@@

\hypertarget{refs}{}
\hypertarget{ref-siegel_cancer_2016}{}
1. Siegel RL, Miller KD, Jemal A. Cancer statistics, 2016. CA: a cancer
journal for clinicians. 2016;66:7--30.

\hypertarget{ref-haggar_colorectal_2009}{}
2. Haggar FA, Boushey RP. Colorectal cancer epidemiology: Incidence,
mortality, survival, and risk factors. Clinics in Colon and Rectal
Surgery. 2009;22:191--7.

\hypertarget{ref-hellinger_reoperation_2006}{}
3. Hellinger MD, Santiago CA. Reoperation for recurrent colorectal
cancer. Clinics in Colon and Rectal Surgery. 2006;19:228--36.

\hypertarget{ref-ryuk_predictive_2014}{}
4. Ryuk JP, Choi G-S, Park JS, Kim HJ, Park SY, Yoon GS, et al.
Predictive factors and the prognosis of recurrence of colorectal cancer
within 2 years after curative resection. Annals of Surgical Treatment
and Research. 2014;86:143--51.

\hypertarget{ref-national_cancer_institute_seer_nodate}{}
5. Institute NC. SEER Cancer Stat Facts: Colon and Rectum Cancer
{[}Internet{]}. {[}cited 2017 Apr 27{]}. Available from:
\url{http://seer.cancer.gov/statfacts/html/colorect.html}

\hypertarget{ref-goodwin_polyamine_2011}{}
6. Goodwin AC, Destefano Shields CE, Wu S, Huso DL, Wu X, Murray-Stewart
TR, et al. Polyamine catabolism contributes to enterotoxigenic
Bacteroides fragilis-induced colon tumorigenesis. Proceedings of the
National Academy of Sciences of the United States of America.
2011;108:15354--9.

\hypertarget{ref-abed_fap2_2016}{}
7. Abed J, Emgård JEM, Zamir G, Faroja M, Almogy G, Grenov A, et al.
Fap2 Mediates Fusobacterium nucleatum Colorectal Adenocarcinoma
Enrichment by Binding to Tumor-Expressed Gal-GalNAc. Cell Host \&
Microbe. 2016;20:215--25.

\hypertarget{ref-arthur_microbial_2014}{}
8. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
McCafferty J, et al. Microbial genomic analysis reveals the essential
role of inflammation in bacteria-induced colorectal cancer. Nature
Communications. 2014;5:4724.

\hypertarget{ref-kostic_fusobacterium_2013}{}
9. Kostic AD, Chun E, Robertson L, Glickman JN, Gallini CA, Michaud M,
et al. Fusobacterium nucleatum potentiates intestinal tumorigenesis and
modulates the tumor-immune microenvironment. Cell Host \& Microbe.
2013;14:207--15.

\hypertarget{ref-wu_human_2009}{}
10. Wu S, Rhee K-J, Albesiano E, Rabizadeh S, Wu X, Yen H-R, et al. A
human colonic commensal promotes colon tumorigenesis via activation of T
helper type 17 T cell responses. Nature Medicine. 2009;15:1016--22.

\hypertarget{ref-zackular_manipulation_2016}{}
11. Zackular JP, Baxter NT, Chen GY, Schloss PD. Manipulation of the Gut
Microbiota Reveals Role in Colon Tumorigenesis. mSphere. 2016;1.

\hypertarget{ref-zackular_gut_2013}{}
12. Zackular JP, Baxter NT, Iverson KD, Sadler WD, Petrosino JF, Chen
GY, et al. The gut microbiome modulates colon tumorigenesis. mBio.
2013;4:e00692--00613.

\hypertarget{ref-baxter_structure_2014}{}
13. Baxter NT, Zackular JP, Chen GY, Schloss PD. Structure of the gut
microbiome following colonization with human feces determines colonic
tumor burden. Microbiome. 2014;2:20.

\hypertarget{ref-flynn_metabolic_2016}{}
14. Flynn KJ, Baxter NT, Schloss PD. Metabolic and Community Synergy of
Oral Bacteria in Colorectal Cancer. mSphere. 2016;1.

\hypertarget{ref-wang_structural_2012}{}
15. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, et al. Structural
segregation of gut microbiota between colorectal cancer patients and
healthy volunteers. The ISME journal. 2012;6:320--9.

\hypertarget{ref-chen_decreased_2013}{}
16. Chen H-M, Yu Y-N, Wang J-L, Lin Y-W, Kong X, Yang C-Q, et al.
Decreased dietary fiber intake and structural alteration of gut
microbiota in patients with advanced colorectal adenoma. The American
Journal of Clinical Nutrition. 2013;97:1044--52.

\hypertarget{ref-chen_human_2012}{}
17. Chen W, Liu F, Ling Z, Tong X, Xiang C. Human intestinal lumen and
mucosa-associated microbiota in patients with colorectal cancer. PloS
One. 2012;7:e39743.

\hypertarget{ref-shen_molecular_2010}{}
18. Shen XJ, Rawls JF, Randall T, Burcal L, Mpande CN, Jenkins N, et al.
Molecular characterization of mucosal adherent bacteria and associations
with colorectal adenomas. Gut Microbes. 2010;1:138--47.
